%% MoCHII — quasi-random (RQMC) photon launching: method, validation, and the
%% diffuse-bin fix.
%% Author: Kwang-Il Seon
%% Written 2026-07-18 for kiseon_memo.cls.
\documentclass[english,a4paper]{kiseon_memo}
\synctex=1
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{longtable}

\newcommand{\code}[1]{\texttt{#1}}

\begin{document}

\title{Quasi-Random Photon Launching in MoCHII:\\
       Implementation, Validation, and the Diffuse-Bin Fix}
\author{Kwang-Il Seon}
\date{\docmoddate}
\maketitle

\begin{abstract}
MoCHII gains an optional randomized quasi-Monte Carlo (RQMC) photon launch:
Owen-scrambled Sobol coordinates replace the Mersenne Twister draws used to
select a packet's source component, frequency bin, direction, and entry
surface, while every post-launch transport event (forced interaction,
Henyey--Greenstein scattering, Russian roulette) stays pseudo-random.  The
Sobol point is addressed by the \emph{global} photon number, so the launch set
is independent of the MPI task count.  On the fixed-state analytic attenuation
gate the cell-wise $\Gamma_{\rm HI}$ error scales as $\sim 1/N$ rather than the
Monte Carlo $1/\sqrt{N}$, giving effective photon gains of $29\times$ at
$2^{17}$ packets and $222\times$ at $2^{20}$, at unchanged wall time and with no
significant bias.  The stage-2 source gates (multiple point sources, external
field) improve in step, and a diffuse-field variant reduces the replicate
scatter of the ionized volume by $1.73\times$.  This work also uncovered and
fixed an unrelated defect in the diffuse-field energy-to-bin mapping that had
become active once threshold-aligned bins (\code{par\%ion\_align\_edges})
became the default: a closed-form log-grid formula mis-binned recombination
photons emitted at an ionization threshold, sending He\,\textsc{i}
ground-recombination photons into the bin below the He\,\textsc{i} edge; the
corrected binary-search mapping lowers the mean neutral-helium fraction on a
diffuse Str\"omgren sphere by $50.5\%$.  The design rationale is recorded in
\code{docs/QUASI\_RANDOM\_LAUNCH.md}.
\end{abstract}

\tableofcontents

%==========================================================================
\section{Motivation and design}
MoCHII's ionizing band has no scattering: given a packet's launch frequency,
position, and direction, the attenuation along its ray is integrated
\emph{analytically} (the zero-variance edge walk).  Consequently the residual
noise in the stellar radiation field comes almost entirely from the sampling of
the launch variables --- which source component, which frequency bin, and which
direction each packet takes --- and \emph{not} from the transport estimator.
This is exactly the situation an RQMC launch is built for: a low-dimensional
integral with a deterministic post-launch response, so a low-discrepancy launch
set directly reduces the noise in global photoionization rates, ionized volume,
and the iteration-to-iteration convergence measures without touching the
transport.

The design is deliberately hybrid.  Only the fixed-dimensional launch step uses
scrambled Sobol coordinates; every conditional, variable-length draw after
launch stays on the Mersenne Twister.  Dust-scattering histories consume a
variable number of uniforms whose meaning depends on the packet's previous
events, so mapping them onto sequential Sobol dimensions would be both unstable
and high-dimensional.  Keeping them pseudo-random preserves the unbiased
scattered field and confines the low-discrepancy structure to the launch cube,
where it helps most.

Two properties of MoCHII make the launch cube favorable.  First, the frequency
variable is a \emph{discrete bin index}: inverting a stratified Sobol coordinate
through the band luminosity CDF automatically delivers near-proportional packet
counts in every bin (for $N = 2^m$ a bin of probability $p$ deviates from $pN$
by order one), removing the binomial bin-count noise that plain Monte Carlo
inherits.  Second, the launch is at most seven-dimensional (nine for the diffuse
field), well within the range where scrambled Sobol nets retain useful variance
reduction even though AMR cell boundaries and ionization fronts make the
integrand only piecewise smooth.  No ideal $N^{-1}(\log N)^d$ rate is assumed;
the gain is measured (Section~\ref{sec:val}).

%==========================================================================
\section{Method}
\label{sec:method}

\subsection{Sobol point by global index}
The generator (\code{src/qmc\_mod.f90}) is self-contained; no external library
is used.  The unscrambled Sobol integer for the 0-based global photon index $i$
and dimension $d$ is the random-access XOR of direction numbers over the set
bits of $i$,
\begin{equation}
  x_d(i) \;=\; \bigoplus_{k\,:\,{\rm bit}\,k\,{\rm of}\,i\,{\rm set}} V_d(k),
\end{equation}
with 32-bit direction numbers $V_d(k)$.  The cost is $O(\mathrm{popcount}(i))$
integer XORs, with no stored state and no sequential dependence, so any packet's
coordinates are computed directly from its global number.  Dimension~1 is the
van der Corput sequence in base~2 ($V_1(k) = 2^{32-k}$); dimensions 2--12 use
the Joe \& Kuo (2008) \code{new-joe-kuo-6} primitive polynomials and initial
direction integers, expanded by the Joe--Kuo recurrence.  \code{QMC\_MAXDIM}${}
= 12$ covers the seven-dimensional stellar superset and the nine-dimensional
diffuse launch.  The unscrambled coordinates reproduce
\code{scipy.stats.qmc.Sobol} (which uses the same table) exactly.

\subsection{Owen scrambling by hashing}
A bare Sobol sequence has no ordinary statistical error estimate and can form
coherent patterns with a Cartesian or octree grid, so the coordinates are
Owen-scrambled.  Full tree-based scrambling is unnecessary; MoCHII uses
hash-based nested-uniform scrambling (Laine \& Karras 2011; Burley 2020):
reverse the 32 bits of $x_d$, apply a keyed Laine--Karras avalanche permutation
(each bit is allowed to affect only less-significant bits), and reverse back.
This is $O(1)$ per coordinate and preserves the net's equidistribution while
producing statistically independent randomized replicates.  The scrambled
integer $k$ maps to the open unit interval as
\begin{equation}
  u \;=\; (k + \tfrac12)\,2^{-32} \;\in\; (0,1),
\end{equation}
so every coordinate is strictly interior, removing a failure class for any later
mapping that takes a logarithm.

Two decorrelated scramble streams share the one Sobol net: the stellar launch
(\code{QMC\_STREAM\_STELLAR}) and the diffuse launch
(\code{QMC\_STREAM\_DIFFUSE}).  A distinct 32-bit key for each dimension and
stream is derived deterministically from \code{par\%qmc\_seed} by an integer
avalanche hash, so the two streams behave as independent replicates while both
remain balanced scrambles of the same net.

\subsection{Fixed dimension layouts}
The meaning of every dimension is fixed and never reassigned conditionally: a
component-dependent reassignment would shift all later coordinates and destroy
the intended low-dimensional projections.  A point component simply leaves its
surface dimensions unused; unused-but-fixed dimensions are harmless because the
projections onto the consumed subset remain those of the scrambled net.

A single internal point source (stage~1) uses the three-dimensional layout of
Table~\ref{tab:dim1}; every other source configuration uses the seven-dimension
superset of Table~\ref{tab:dim7}; diffuse recombination packets use the
nine-dimension layout of Table~\ref{tab:dim9} on the separate diffuse stream.

\begin{table}[htbp]
\centering
\small
\caption{Single internal point source (\code{QMC\_NDIM\_USED}${}=3$).}
\label{tab:dim1}
\begin{tabular}{cll}
\toprule
dim & variable & mapping \\
\midrule
1 & frequency bin & discrete band luminosity CDF \\
2 & polar direction & $\mu = 2u-1$ \\
3 & azimuth & $\phi = 2\pi u$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\small
\caption{Superset layout for every non-single-point configuration
(multiple point sources, external field, or a mixed set).}
\label{tab:dim7}
\begin{tabular}{cll}
\toprule
dim & variable & mapping \\
\midrule
1 & source component & component luminosity CDF \\
2 & frequency bin & that component's spectrum CDF \\
3 & polar / incidence & point: $\mu = 2u-1$; \ external: Lambert $\mu = \sqrt{u}$ \\
4 & azimuth & $\phi = 2\pi u$ \\
5 & entry surface & rectangular entry face / sphere entry $\cos\theta$ \\
6 & first surface coordinate & in-face coordinate / sphere entry azimuth \\
7 & second surface coordinate & in-face coordinate (unused for the sphere) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\small
\caption{Diffuse recombination packets (\code{par\%diffuse\_field}), on the
diffuse stream.}
\label{tab:dim9}
\begin{tabular}{cll}
\toprule
dim & variable & mapping \\
\midrule
1 & emitting leaf & leaf luminosity CDF (binary search) \\
2--4 & position in leaf & uniform, $2u-1$ per axis \\
5 & polar direction & $\mu = 2u-1$ \\
6 & azimuth & $\phi = 2\pi u$ \\
7 & emission channel & channel CDF (H\,\textsc{ii}/He\,\textsc{ii}/He\,\textsc{iii}/He\,\textsc{i}) \\
8 & first energy variate & ${\rm Exp}(1)$ for the ground channels; He\,\textsc{i} branch pick \\
9 & second energy variate & He\,\textsc{i} two-photon flat energy (unused otherwise) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Common random numbers and replicates}
Using the same scrambled launch set at every ionization/thermal iteration is a
form of common random numbers: it correlates the radiation estimates before and
after an opacity update and reduces noise in their difference, which stabilizes
the convergence measures and separates front motion from sampling noise.  The
scramble key is \code{par\%qmc\_seed}, held fixed through all nonlinear
iterations of one solve.  A fixed scramble can, however, imprint a repeatable
quadrature pattern that the nonlinear solution adapts to, making iteration
differences look quieter than the true sampling uncertainty; convergence under
one scramble is therefore not a substitute for replicate agreement.  For a
production error estimate, run independent replicates with different
\code{qmc\_seed} values --- at least four, more when cost permits --- and take
their scatter as the RQMC uncertainty.

%==========================================================================
\section{Implementation}
\label{sec:impl}
The generator lives in \code{src/qmc\_mod.f90}; the launch mappings that consume
its coordinates are \code{gen\_ion\_photon\_qmc} in \code{src/ion\_band\_mod.f90}
(stellar) and \code{gen\_diffuse\_photon\_qmc} in \code{src/diffuse\_mod.f90}
(diffuse).  Two parameters control the option:
\begin{itemize}
\item \code{par\%launch\_sequence} = \code{'random'} (default; the Mersenne
  Twister MT19937-64 path, arithmetically unchanged) or \code{'sobol'};
\item \code{par\%qmc\_seed} (default \code{12345}) --- the scramble/replicate
  key, fixed across the gas iterations.
\end{itemize}
The stellar loop in \code{src/main.f90} addresses the Sobol point by the global
photon number $i-1$ (0-based), and the diffuse loop does the same on its own
stream indexed by the global diffuse photon number.  Because the coordinates
depend only on the global index and not on which rank handles a packet, the
launch \emph{set} is identical for any MPI task count; static or strided work
assignment does not change the sample.

Everything after launch stays pseudo-random: forced first interaction,
Henyey--Greenstein direction sampling, subsequent optical depths, and Russian
roulette all draw from the rank-local Mersenne Twister exactly as before.
Bitwise MPI-count independence of a \emph{full} run is therefore obtained only
in absorption-only transport, where the post-launch path is deterministic; with
dust scattering enabled the launch set is still reproducible but the scattered
histories are not.  The multi-wavelength dust/SED launch (\code{src/sed\_mod.f90})
is not exercised by the ionizing workflow and is left on its alias sampler; its
inverse-CDF conversion is documented as unnecessary for now.

%==========================================================================
\section{The diffuse bin-mapping fix}
\label{sec:binfix}
This work uncovered a defect independent of QMC, in the diffuse-field packet
generator, and fixes it.

\paragraph{The bug.}  \code{gen\_diffuse\_photon} mapped a sampled photon energy
to its band bin with a closed-form expression that assumed the legacy
log-uniform bin grid.  Once threshold-aligned bins
(\code{par\%ion\_align\_edges}) became the default, the bin edges sit \emph{on}
the ionization thresholds, and the closed form mis-assigns an energy at a
threshold: a $24.587$~eV He\,\textsc{i} ground-recombination photon landed in
the bin \emph{below} the He\,\textsc{i} edge, where it could never ionize
helium.

\paragraph{The fix.}  The band edge array is now stored once
(\code{ion\_eedge}, valid for both grids), and \code{ion\_bin\_of(eph)} finds
the bin by binary search: \code{ion\_eedge}$(b) \le E_{\rm ph} <
{}$\code{ion\_eedge}$(b{+}1)$, so an energy exactly on a threshold edge falls in
the bin \emph{above} it --- the physically correct side for a recombination
photon emitted at threshold${}+\delta$.  Both the pseudo-random and the QMC
diffuse generators call \code{ion\_bin\_of}.

\paragraph{Measured impact.}  On a converged diffuse-field Str\"omgren sphere on
the aligned (default) grid, buggy versus fixed: the mean $x_{\rm HeI}$ drops
$50.5\%$ ($0.0147 \to 0.0073$), the mean $x_{\rm HeII}$ rises $0.75\%$, and
$V_{\rm ion}$/$R_{\rm eff}$ are unchanged (hydrogen is stellar-dominated).  Unit
checks: on the legacy log grid the binary search reproduces the closed form over
a two-million-point energy scan except at a single exact-edge floating-point
tie; on the aligned grid the H\,\textsc{i}/He\,\textsc{i}/He\,\textsc{ii}
thresholds are bin edges and threshold${}+\delta$ lands in the bin above.  Runs
with \code{diffuse\_field} off are arithmetically untouched, and the recorded
diffuse validations in the project log predate the aligned-edge default, so the
bug had contaminated no recorded gate.

%==========================================================================
\section{Validation}
\label{sec:val}
Every number below is measured and independently verified.

\subsection{Generator gates (Q0/Q2)}
The unscrambled Sobol coordinates for dimensions 1--12 match
\code{scipy.stats.qmc.Sobol(d=12, scramble=False)} exactly (sorted integer
lattices, maximum code difference 0).  The Owen-scrambled sets are strictly
interior; the one-dimensional dyadic counts are exact at every level; the
dimension-$(1,2)$ $(0,m,2)$-net and the elementary-interval occupancy are
preserved under scrambling, for seeds \code{12345}/\code{777}/\code{2024} and
both streams.  Different seeds give different but individually balanced nets
(maximum coordinate shift $|\Delta u| = 0.84$--$0.98$), and the stellar and
diffuse streams differ like independent replicates.

\subsection{Default-path regression}
With \code{launch\_sequence='random'}, at one MPI task under
\code{-fp-model precise}, the pre-change and post-change builds are
\emph{bit-identical} for single-source, multi-source, and external inputs
(worst $|\Delta| = 0$).  The diffuse log-grid input sits at the code-generation
rounding floor (maximum relative difference $6.8\times10^{-16}$, with the bins
proven identical).  At more than one task the difference equals the MPI
\code{ALLREDUCE} reordering floor ($\sim 1$ ULP, the same as running one binary
twice).

\subsection{MPI-count independence}
The Sobol launch set is task-count independent: \code{'sobol'} rates at 4 tasks
versus 8 tasks agree to $6.0\times10^{-15}$ (stage~1) and
$2.5\times10^{-14}$ (stage~2) relative, whereas the \code{'random'} contrast
differs at order unity (each rank owns a different Mersenne Twister stream).

\subsection{Variance gate: analytic attenuation}
The cleanest test is the fixed-state G0 gate (\code{tests/g0\_gamma}):
$\Gamma_{\rm HI}$ against the exact analytic attenuation, comparing four ordinary
Monte Carlo replicates with four RQMC replicates.  Table~\ref{tab:var} reports
the cell-wise and radial-profile RMS error and the effective photon-number gain
(the number of Monte Carlo packets needed to reach the RQMC RMS).

\begin{table}[htbp]
\centering
\small
\caption{Fixed-state $\Gamma_{\rm HI}$ error, Monte Carlo (MC) versus RQMC,
four replicates each.  ``Gain'' is the effective photon-number gain.}
\label{tab:var}
\begin{tabular}{lrrr}
\toprule
 & MC RMS & RQMC RMS & gain \\
\midrule
\multicolumn{4}{@{}l}{\itshape $N = 2^{17}$ packets}\\
cell-wise      & 16.9\% & 3.1\%  & $28.8\times$ \\
radial profile & 0.58\% & 0.13\% & $20.4\times$ \\
\addlinespace
\multicolumn{4}{@{}l}{\itshape $N = 2^{20}$ packets}\\
cell-wise      & 6.0\%  & 0.40\% & $222\times$ \\
radial profile & 0.24\% & 0.12\% & $4.1\times$ \\
\bottomrule
\end{tabular}
\end{table}

The RQMC cell-wise error scales as $\sim 1/N$ (Monte Carlo $\sim 1/\sqrt{N}$),
so the gain grows with $N$.  The RQMC bias is $\sim 0.06\%$, stable, and smaller
than the Monte Carlo bias ($0.12$--$0.41\%$): no significant bias.  One caveat is
reported honestly: at $2^{20}$ the RQMC radial-profile error saturates at a
systematic floor (all four replicates give an identical $0.119\%$, with a fixed
$-0.06\%$ bias) that is band-discretization systematics, not sampling noise, so
the $4.1\times$ profile gain there is a lower bound.  Sequence generation is a
few tens of integer operations per packet, negligible against ray traversal, so
the wall time is unchanged.

\subsection{Full-physics Str\"omgren smoke}
On a full-physics Str\"omgren run, $R_{\rm eff}$ under \code{'sobol'} differs
from \code{'random'} by $-0.039\%$ (within noise).  The launch-noise reduction
shows up directly in the convergence measure: the \code{'sobol'} run reached the
volume-integrated criterion in 57 iterations where the \code{'random'} run hit
the 60-iteration cap unconverged.  The $x_{\rm HI}$ slices are clean --- no
angular spokes and no grid-aligned artifacts (inspected).

\subsection{Stage-2 source gates}
The existing stage-2 analytic checkers, rerun under \code{'sobol'} at their
recorded criteria, improve or match the \code{'random'} numbers
(Table~\ref{tab:stage2}).

\begin{table}[htbp]
\centering
\small
\caption{Stage-2 source gates under \code{'sobol'} versus the recorded
\code{'random'} values.}
\label{tab:stage2}
\begin{tabular}{lll}
\toprule
gate & \code{'sobol'} & \code{'random'} \\
\midrule
two-point superposition (median) & 0.14\% & 0.76\% \\
two-point superposition (mean ratio) & 0.9989 & --- \\
external rec, interior $\langle J\rangle/J$ & 0.9984 & 0.9981 \\
external sph, interior $\langle J\rangle/J$ & 0.9990 & 0.9991 \\
mixed point${}+{}$external (median) & 0.13\% & 0.75\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Diffuse variance}
On a diffuse-field Str\"omgren sphere from a neutral start, the replicate scatter
(standard deviation over three seeds) of $V_{\rm ion}$ and $R_{\rm eff}$ is
$1.73\times$ smaller for RQMC than for Monte Carlo, and unbiased.  This is a
smaller gain than the stellar launch, as the design anticipated: the emitting-leaf
CDF is strongly discontinuous and changes as the gas state iterates, and the
diffuse cube is higher-dimensional.

%==========================================================================
\section{Usage guidance and limitations}
Set \code{par\%launch\_sequence = 'sobol'} to enable the RQMC launch and prefer a
power-of-two photon count, \code{nphotons}${} = 2^m$, where the Sobol net has its
cleanest balance.  Keep \code{par\%qmc\_seed} fixed across the nonlinear
iterations of one solve (common random numbers), and change it only to generate
independent replicates; take the scatter of at least four replicates as the RQMC
error estimate.  The default remains \code{'random'} so recorded gates reproduce.

The gains follow the design expectation.  The largest reductions are for global,
direct-field quantities: direct stellar photoionization rates, integrated
ionized volume and effective Str\"omgren radius, volume-integrated convergence
measures, smooth radial profiles, and early iterations run with a modest photon
budget.  Smaller or inconsistent gains are expected for the maximum error in a
single I-front cell, narrow shadow boundaries and clump skins, highly localized
line emissivities, high-resolution image pixels, scattering-dominated paths, and
diffuse-field-dominated cells.  Low discrepancy does not remove physical
discretization errors --- frequency binning, AMR resolution, atomic-data fits,
and nonlinear stopping tolerances remain independent error sources --- and no
universal speedup factor should be assumed; a useful reduction is best reported
as an effective photon-number gain.  By design, variable-length scattering
histories remain on the Mersenne Twister; extending RQMC into them would need a
maximum-order dimension policy and a separate validation campaign.

%==========================================================================
\section*{References}
\addcontentsline{toc}{section}{References}
\begin{itemize}
\item Sobol', I. M. 1967, \textit{USSR Comput.\ Math.\ Math.\ Phys.}, 7, 86.
\item Owen, A. B. 1997, \textit{SIAM J.\ Numer.\ Anal.}, 34, 1884;
      1998, \textit{ACM Trans.\ Model.\ Comput.\ Simul.}, 8, 71
      (nested-uniform scrambling).
\item Joe, S., \& Kuo, F. Y. 2008, \textit{SIAM J.\ Sci.\ Comput.}, 30, 2635
      (\code{new-joe-kuo-6} direction numbers).
\item Laine, S., \& Karras, T. 2011, in \textit{Proc.\ High-Performance
      Graphics} (stratified sampling for stochastic transparency).
\item Burley, B. 2020, ``Practical Hash-based Owen Scrambling,''
      \textit{J.\ Comput.\ Graph.\ Tech.}, 9, 1.
\item Seon, K.-I. 2026, \code{docs/QUASI\_RANDOM\_LAUNCH.md} (MoCHII design
      document).
\end{itemize}

\end{document}
